Apparent preferences for cognitive effort fade when multiple forms of effort

### 2020-12-04 Cost_manuscript.Rmd
commenting out old modeling approach, and adding intro paragraph to modeling results

``` {r BTW: 16.3.3. OC modeling, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
## fit the OC model
summaryOC_btw <- list()
summaryOC_btw$all <- dataBtw %>%
                 group_by(Cost, SubjID) %>%
                 do(optimizeOCModel(., simplify = T)) %>%
                 ungroup()

## plot
summaryOC_btw$plot <- ggplot(summaryOC_btw$all, aes(Cost, Gamma, fill = Cost)) +
  # geom_hline(yintercept = 0.64, alpha = 0.9, color = "gray20", size = 1) +
  # geom_hline(yintercept = 0.7, alpha = 0.9, color = "gray40", size = 1) +
  # geom_hline(yintercept = 0.9, alpha = 0.9, color = "gray60", size = 1) +
  geom_hline(yintercept = 0.74, alpha = 0.9, size = 1, linetype = "dashed") + # mean of highest earning rates across blocks
  #geom_dotplot(stackdir = 'center', binaxis = 'y', show.legend = F) +
  geom_boxplot(show.legend = F) +
  geom_jitter(width = 0.1, alpha = 0.5, show.legend = F, pch = 21, size = 3) +
  ylim(0,1.5) +
  labs(x = "") +
  scale_fill_manual(values = colsBtw) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"),
        text = element_text(size = 16))

## one sample t against optimal strategy
summaryOC_btw$t.test <- summaryOC_btw$all %>%
  select(Cost, Gamma) %>%
  plyr::dlply("Cost", identity) %>%
  sapply(., function(data) {t.test(data$Gamma, mu = 0.74)})

## descriptive stats
summaryOC_btw$descriptives <- summaryOC_btw$all %>%
  group_by(Cost) %>%
  summarise(mGamma = round(mean(Gamma), digits = 2),
            sdGamma = round(sd(Gamma), digits = 2))

@@

``` {r BTW: 16.3.5. OC comparisons, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# ANOVA
OCanova <- list()
OCanova$aov <- summary(aov(Gamma ~ Cost, data = summaryOC_btw$all))
Fval <- round(OCanova$aov[[1]]$`F value`[1], digits = 2) # doing separate variables because I can't easily use the `` in markdown.
Pval <- OCanova$aov[[1]]$`Pr(>F)`[1]
Rsquared_aov <- round(OCanova$aov[[1]]$`Sum Sq`[1] / sum(OCanova$aov[[1]]$`Sum Sq`), digits = 2)

# Post-hoc pairwise tests
OCanova$perms <- summaryOC_btw$all %>%
  select(Cost, Gamma) %>%
  plyr::dlply("Cost", identity) %>%
  do.call(cbind, .) %>%
  select_if(is.numeric) %>%
  rename(Physical = Physical.Gamma,
         Cognitive = Cognitive.Gamma,
         Wait = Wait.Gamma,
         Easy = Easy.Gamma) %>%
  outer(., ., FUN = Vectorize(permute), simple = T)
diag(OCanova$perms) <- 1

OCanova$ES <- summaryOC_btw$all %>%
  select(Cost, Gamma) %>%
  plyr::dlply("Cost", identity) %>%
  do.call(cbind, .) %>%
  select_if(is.numeric) %>%
  rename(Physical = Physical.Gamma,
         Cognitive = Cognitive.Gamma,
         Wait = Wait.Gamma,
         Easy = Easy.Gamma) %>%
  outer(., ., FUN = Vectorize(cohenD)) %>%
  abs()
```

``` {r BTW: 16.3.5. corrplot, echo = FALSE, fig.align = "center", fig.width = 5, fig.height = 5}
# # plot all pairwise comparisons
# corrplot(OCanova$ES,
#          is.corr = F, 
#          p.mat = OCanova$perms,
#          type = "upper",
#          insig = "p-value",
#          sig.level = -1,
#          na.label = "square",
#          na.label.col = "grey",
#          method = "color", 
#          tl.col = "black", 
#          tl.cex = 0.8,
#          outline = T)
```

``` {r BTW: 16.3.4. OC cross-validation, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# Compute the OC for pre-midpoint
summaryOC_btw$pre <- dataBtw %>%
                 filter(Half == "Half_1") %>%
                 group_by(Cost, SubjID) %>%
                 do(optimizeOCModel(., simplify = T)) %>%
                 ungroup() %>%
                 select(SubjID, Gamma)

# let's add the gammas to the second half data and estimate prop predicted on the second half
summaryOC_btw$predicted <- dataBtw %>%
                 filter(Half == "Half_2") %>%
                 group_by(Cost, SubjID) %>%
                 left_join(summaryOC_btw$pre, by = "SubjID") %>%
                 mutate(OC = Handling * Gamma,
                        predicted = Offer > OC) %>% 
                 summarise(percentPredicted = mean(predicted == Choice) * 100) %>%
                 group_by(Cost) %>%
                 summarise(mPredicted = round(mean(percentPredicted), digits = 2),
                           sdPredicted = round(sd(percentPredicted), digits = 2))

@@

  We tested three nested models based on a set of criteria derived from observed choice behavior (see Methods for details). The first one, a base formulation, estimated a subjective opportunity cost for each individual and demand. The second one allowed the OC to gradually adapt using a reinforcement learning rule. The final one added a time sensitivity parameter that controlled how each demand affected the perceived length of the handling time, which could be recalibrated based on the experience of multiple types of demands.

@@

``` {r WTH: 16.3.4., echo = FALSE, fig.align="center", fig.width=6, fig.height=4}
# Here it might be best to estimate gamma for the wait trials only at first
# once the wait OCs are estimated, then they can be modulated by effort
# Estimate gamma per subject for each group separately
summaryOC_wth <- list()
summaryOC_wth$all <- dataWth %>%
  mutate(Block = case_when(
    Block %in% c(1, 2) ~ 1,
    Block %in% c(3, 4) ~ 2,
    Block %in% c(5, 6) ~ 3
  )) %>%
  #filter(Choice < 2) %>%
  group_by(SubjID, Cost) %>% #group_by(SubjID, Cost, Block) %>%
  do(optimizeOCModel(., simplify = T)) %>%
  ungroup()

# plot
summaryOC_wth$plot <- ggplot(summaryOC_wth$all, aes(Cost, Gamma, fill = Cost)) +
    geom_boxplot(show.legend = F) +
    geom_hline(yintercept = 0.64, alpha = 0.9, color = "gray20", size = 1) +
    geom_hline(yintercept = 0.7, alpha = 0.9, color = "gray40", size = 1) +
    geom_hline(yintercept = 0.9, alpha = 0.9, color = "gray60", size = 1) +
    ylim(0,1.5) +
    labs(x = "") +
    scale_fill_manual(values = colsWth) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black"),
          text = element_text(size = 16))

``` {r WTH: 16.3.5., echo = FALSE, fig.width=5, fig.height=4, fig.align="center"}
# subdivide data
dataTrain <- dataWth %>%
   filter(Block < 5)

dataTest <- dataWth %>%
  filter(Block > 4)

# estimate
# per-cost, but consider the idea above
OCs_validation <- list()
temp <- dataTrain %>% filter(Cost == "Wait-C") %>% plyr::dlply("SubjID", identity)
OCs_validation$`Wait-C` <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
temp <- dataTrain %>% filter(Cost == "Wait-P") %>% plyr::dlply("SubjID", identity)
OCs_validation$`Wait-P` <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
temp <- dataTrain %>% filter(Cost == "Cognitive") %>% plyr::dlply("SubjID", identity)
OCs_validation$Cognitive <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
temp <- dataTrain %>% filter(Cost == "Physical") %>% plyr::dlply("SubjID", identity)
OCs_validation$Physical <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})

# reshape data
temp <- bind_rows(OCs_validation) %>% 
  mutate(SubjID = subjList_wth) %>%
  gather(Cost, gamma, `Wait-C`:Physical) %>%
  left_join(dataTest, ., by = c("SubjID", "Cost")) %>% # assign OCs to each subject per cost type
  mutate(prediction = Offer > Handling * gamma) %>%
  # plyr::dlply("SubjID", identity)
  group_by(SubjID, Cost) %>%
  summarize(predicted = mean(Choice == prediction))

``` {r WTH: 16.3.6., echo = FALSE}
# repeated measures anova
OCs_anova <- list()
OCs_anova$aov <- summary(aov(Gamma ~ Cost + Error(SubjID / Cost), data = summaryOC_wth$all))

# table
#pander(OCs_anova$aov)

# do all pairwise comparisons
OCs_anova$gammas <- summaryOC_wth$all %>% 
  reshape2::dcast(SubjID ~ Cost, value.var = "Gamma") %>%
  select(-SubjID)

# perform a paired permutation among every pair of cost types
OCs_anova$perms <- outer(OCs_anova$gammas, OCs_anova$gammas, FUN = Vectorize(permute), simple = T, paired = T)
OCs_anova$perms[rbind(c(1,4),c(2,3),c(3,2),c(4,1))] <- NA

# and the effect sizes
OCs_anova$ES <- abs(outer(OCs_anova$gammas, OCs_anova$gammas, FUN = Vectorize(DescTools::CohenD)))
OCs_anova$ES[rbind(c(1,4),c(2,3),c(3,2),c(4,1))] <- NA
```

``` {r WTH: 16.3.6. corrplot, echo = FALSE, fig.align="center", fig.width=5, fig.height=5}
# corrplot(OCs_anova$ES,
#          is.corr = F, 
#          p.mat = OCs_anova$perms,
#          type = "upper",
#          insig = "p-value",
#          sig.level = -1,
#          na.label = "square",
#          na.label.col = "grey",
#          method = "color", 
#          tl.col = "black", 
#          tl.cex = 0.8,
#          outline = T)
```

>>>>

``` {r OLD BTW and WTH: 16.3. OC modeling btw, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# ## fit the OC model
# summaryOC_btw <- list()
# summaryOC_btw$all <- dataBtw %>%
#                  group_by(Cost, SubjID) %>%
#                  do(optimizeOCModel(., simplify = T)) %>%
#                  ungroup()
# 
# 
# ## plot
# summaryOC_btw$plot <- ggplot(summaryOC_btw$all, aes(Cost, Gamma, fill = Cost)) +
#   # geom_hline(yintercept = 0.64, alpha = 0.9, color = "gray20", size = 1) +
#   # geom_hline(yintercept = 0.7, alpha = 0.9, color = "gray40", size = 1) +
#   # geom_hline(yintercept = 0.9, alpha = 0.9, color = "gray60", size = 1) +
#   geom_hline(yintercept = 0.74, alpha = 0.9, size = 1, linetype = "dashed") + # mean of highest earning rates across blocks
#   #geom_dotplot(stackdir = 'center', binaxis = 'y', show.legend = F) +
#   geom_boxplot(show.legend = F) +
#   geom_jitter(width = 0.1, alpha = 0.5, show.legend = F, pch = 21, size = 3) +
#   ylim(0,1.5) +
#   labs(x = "") +
#   scale_fill_manual(values = colsBtw) +
#   theme(panel.grid.major = element_blank(),
#         panel.grid.minor = element_blank(),
#         panel.background = element_blank(),
#         axis.line = element_line(colour = "black"),
#         text = element_text(size = 16))
# 
# 
# ## one sample t against optimal strategy
# summaryOC_btw$t.test <- summaryOC_btw$all %>%
#   select(Cost, Gamma) %>%
#   plyr::dlply("Cost", identity) %>%
#   sapply(., function(data) {t.test(data$Gamma, mu = 0.74)})
# 
# 
# ## descriptive stats
# summaryOC_btw$descriptives <- summaryOC_btw$all %>%
#   group_by(Cost) %>%
#   summarise(mGamma = round(mean(Gamma), digits = 2),
#             sdGamma = round(sd(Gamma), digits = 2))

@@

``` {r OLD BTW and WTH: 16.3. OC comparisons btw, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# # ANOVA
# OCanova <- list()
# OCanova$aov <- summary(aov(Gamma ~ Cost, data = summaryOC_btw$all))
# Fval <- round(OCanova$aov[[1]]$`F value`[1], digits = 2) # doing separate variables because I can't easily use the `` in markdown.
# Pval <- OCanova$aov[[1]]$`Pr(>F)`[1]
# Rsquared_aov <- round(OCanova$aov[[1]]$`Sum Sq`[1] / sum(OCanova$aov[[1]]$`Sum Sq`), digits = 2)
#   
# # Post-hoc pairwise tests
# OCanova$perms <- summaryOC_btw$all %>%
#   select(Cost, Gamma) %>%
#   plyr::dlply("Cost", identity) %>%
#   do.call(cbind, .) %>%
#   select_if(is.numeric) %>%
#   rename(Physical = Physical.Gamma,
#          Cognitive = Cognitive.Gamma,
#          Wait = Wait.Gamma,
#          Easy = Easy.Gamma) %>%
#   outer(., ., FUN = Vectorize(permute), simple = T)
# diag(OCanova$perms) <- 1
# 
# OCanova$ES <- summaryOC_btw$all %>%
#   select(Cost, Gamma) %>%
#   plyr::dlply("Cost", identity) %>%
#   do.call(cbind, .) %>%
#   select_if(is.numeric) %>%
#   rename(Physical = Physical.Gamma,
#          Cognitive = Cognitive.Gamma,
#          Wait = Wait.Gamma,
#          Easy = Easy.Gamma) %>%
#   outer(., ., FUN = Vectorize(cohenD)) %>%
#   abs()
```

``` {r OLD BTW and WTH: 16.3. corrplot btw, echo = FALSE, fig.align = "center", fig.width = 5, fig.height = 5}
# # plot all pairwise comparisons
# corrplot(OCanova$ES,
#          is.corr = F, 
#          p.mat = OCanova$perms,
#          type = "upper",
#          insig = "p-value",
#          sig.level = -1,
#          na.label = "square",
#          na.label.col = "grey",
#          method = "color", 
#          tl.col = "black", 
#          tl.cex = 0.8,
#          outline = T)
```

``` {r OLD BTW and WTH: 16.3. OC cross-validation btw, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# # Compute the OC for pre-midpoint
# summaryOC_btw$pre <- dataBtw %>%
#                  filter(Half == "Half_1") %>%
#                  group_by(Cost, SubjID) %>%
#                  do(optimizeOCModel(., simplify = T)) %>%
#                  ungroup() %>%
#                  select(SubjID, Gamma)
# 
# # let's add the gammas to the second half data and estimate prop predicted on the second half
# summaryOC_btw$predicted <- dataBtw %>%
#                  filter(Half == "Half_2") %>%
#                  group_by(Cost, SubjID) %>%
#                  left_join(summaryOC_btw$pre, by = "SubjID") %>%
#                  mutate(OC = Handling * Gamma,
#                         predicted = Offer > OC) %>% 
#                  summarise(percentPredicted = mean(predicted == Choice) * 100) %>%
#                  group_by(Cost) %>%
#                  summarise(mPredicted = round(mean(percentPredicted), digits = 2),
#                            sdPredicted = round(sd(percentPredicted), digits = 2))

@@

``` {r OLD BTW and WTH: 16.3. OC Modeling wth, echo = FALSE, fig.align="center", fig.width=6, fig.height=4}
# # Here it might be best to estimate gamma for the wait trials only at first
# # once the wait OCs are estimated, then they can be modulated by effort
# # Estimate gamma per subject for each group separately
# summaryOC_wth <- list()
# summaryOC_wth$all <- dataWth %>%
#   mutate(Block = case_when(
#     Block %in% c(1, 2) ~ 1,
#     Block %in% c(3, 4) ~ 2,
#     Block %in% c(5, 6) ~ 3
#   )) %>%
#   #filter(Choice < 2) %>%
#   group_by(SubjID, Cost) %>% #group_by(SubjID, Cost, Block) %>%
#   do(optimizeOCModel(., simplify = T)) %>%
#   ungroup()
# 
# # plot
# summaryOC_wth$plot <- ggplot(summaryOC_wth$all, aes(Cost, Gamma, fill = Cost)) +
#     geom_boxplot(show.legend = F) +
#     geom_hline(yintercept = 0.64, alpha = 0.9, color = "gray20", size = 1) +
#     geom_hline(yintercept = 0.7, alpha = 0.9, color = "gray40", size = 1) +
#     geom_hline(yintercept = 0.9, alpha = 0.9, color = "gray60", size = 1) +
#     ylim(0,1.5) +
#     labs(x = "") +
#     scale_fill_manual(values = colsWth) +
#     theme(panel.grid.major = element_blank(),
#           panel.grid.minor = element_blank(),
#           panel.background = element_blank(),
#           axis.line = element_line(colour = "black"),
#           text = element_text(size = 16))

``` {r OLD BTW and WTH: 16.3. cross-v wth, echo = FALSE, fig.width=5, fig.height=4, fig.align="center"}
# # subdivide data
# dataTrain <- dataWth %>%
#    filter(Block < 5)
# 
# dataTest <- dataWth %>%
#   filter(Block > 4)
# 
# # estimate
# # per-cost, but consider the idea above
# OCs_validation <- list()
# temp <- dataTrain %>% filter(Cost == "Wait-C") %>% plyr::dlply("SubjID", identity)
# OCs_validation$`Wait-C` <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
# temp <- dataTrain %>% filter(Cost == "Wait-P") %>% plyr::dlply("SubjID", identity)
# OCs_validation$`Wait-P` <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
# temp <- dataTrain %>% filter(Cost == "Cognitive") %>% plyr::dlply("SubjID", identity)
# OCs_validation$Cognitive <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
# temp <- dataTrain %>% filter(Cost == "Physical") %>% plyr::dlply("SubjID", identity)
# OCs_validation$Physical <- sapply(temp, function(Data) {optimizeOCModel(Data)$Gamma})
# 
# # reshape data
# temp <- bind_rows(OCs_validation) %>% 
#   mutate(SubjID = subjList_wth) %>%
#   gather(Cost, gamma, `Wait-C`:Physical) %>%
#   left_join(dataTest, ., by = c("SubjID", "Cost")) %>% # assign OCs to each subject per cost type
#   mutate(prediction = Offer > Handling * gamma) %>%
#   # plyr::dlply("SubjID", identity)
#   group_by(SubjID, Cost) %>%
#   summarize(predicted = mean(Choice == prediction))

``` {r OLD BTW and WTH: 16.3. param compare wth, echo = FALSE}
# # repeated measures anova
# OCs_anova <- list()
# OCs_anova$aov <- summary(aov(Gamma ~ Cost + Error(SubjID / Cost), data = summaryOC_wth$all))
# 
# # table
# #pander(OCs_anova$aov)
# 
# # do all pairwise comparisons
# OCs_anova$gammas <- summaryOC_wth$all %>% 
#   reshape2::dcast(SubjID ~ Cost, value.var = "Gamma") %>%
#   select(-SubjID)
# 
# # perform a paired permutation among every pair of cost types
# OCs_anova$perms <- outer(OCs_anova$gammas, OCs_anova$gammas, FUN = Vectorize(permute), simple = T, paired = T)
# OCs_anova$perms[rbind(c(1,4),c(2,3),c(3,2),c(4,1))] <- NA
# 
# # and the effect sizes
# OCs_anova$ES <- abs(outer(OCs_anova$gammas, OCs_anova$gammas, FUN = Vectorize(DescTools::CohenD)))
# OCs_anova$ES[rbind(c(1,4),c(2,3),c(3,2),c(4,1))] <- NA
```

``` {r OLD BTW and WTH: 16.3. corrplot wth, echo = FALSE, fig.align="center", fig.width=5, fig.height=5}
# corrplot(OCs_anova$ES,
#          is.corr = F, 
#          p.mat = OCs_anova$perms,
#          type = "upper",
#          insig = "p-value",
#          sig.level = -1,
#          na.label = "square",
#          na.label.col = "grey",
#          method = "color", 
#          tl.col = "black", 
#          tl.cex = 0.8,
#          outline = T)
```

#summaryOC_btw$plot

``` {r BTW and WTH: 16.3. Model fits, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
# load results from models
load("models_gk/results/btw.Rdata")
load("models_gk/results/wth.Rdata")
```

  We tested four models based on a set of criteria derived from observed choice behavior (see Methods for details). The first one, a base formulation, probed the predictions of the Marginal Value Theorem (Charnov, 1975). The second one allowed the opportunity cost to gradually adapt using a reinforcement learning rule. The third model assumed that different demands imposed distinct biases on the perceived opportunity costs, and that these biases could be recalibrated with experience. A final model combined the adaptive features of the second and third formulations.

@@
